Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-05-10 pkg/glmertree/paper/treatment_interactions.Rnw
Paper revised
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@847 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\linespread{1.2}
\usepackage{tabto}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB without random-effects estimation. GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (in 90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB without random effects (average predictive accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}. Specifically, this model class encompasses the generalized linear model (GLM). GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies. Also, \citeA{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.

In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure. However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \cite<which is not done in other tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Application section, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}. But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.

@@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM is often used to estimate treatment effects. GLMs allow for the choice of a suitable response distribution -- for example normal, binomial, or Poisson - depending on whether the treatment outcome variable is continuous (e.g., an improvement score), binary (e.g., improved or not), or a count (e.g., number of events in a certain time span), respectively. In all cases the expectation $\mu_i$ of the outcome variable $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and a suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function\footnote{An overview of notation used is provided in the appendix.}. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. In case of a continuous response variable, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

@@

In RCTs, treatment effects are often estimated after controlling posttreatment values on the outcome measure for the linear effect of pretreatment values on the same measure. Therefore, we included the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores as an offset variable in growing the GLM and GLMM trees. An offset variable is a linear predictor with an a-priori specified coefficient of one. Including the predictions of a linear regression of HAM-D posttreatment on HAM-D pretreatment scores as an offset has the same effect as statistically controlling for the linear effects of pretreatment scores, as is often done in ANCOVA.

We built all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score. As a result, data from 694 patients from 7 studies were included for the analyses. Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.

Predictive accuracy of GLM and GLMM trees was assessed by calculating the average correlations between observed and predicted HAM-D scores, based on 50-fold cross validation.

<<depression-trees>>=
hrs <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))

## GLM tree
lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, offset = hrs)

## GLMM tree (estimated with joint = FALSE because fixed effect HRSDt0 is included in random formula)
lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(hrs) | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, ranefstart = hrs, joint = F)
@

@@

Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W}. Note that the GLM tree in Figure~\ref{fig:lmtree_C&W} is also the tree in the first iteration of the GLMM tree algorithm.

@@

The GLM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education. Node 2 of Figure~\ref{fig:lmtree_C&W} indicates that for patients with a low level of education, antidepressant medication provides the greatest reduction in HAM-D scores. Node 4 indicates that for patients with a higher level of education, and no comorbid anxiety disorder, the reduction in HAM-D scores is about the same for CBT and PHA. Node 5 indicates, that for patients with a higher level of education and a comorbid anxiety disorder, the reduction in HAM-D scores is greatest for PHA.

@@

By taking into account study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split. The GLMM tree selected only presence of a comorbid anxiety disorder as a partitioning variable. The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores, whereas for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores. The estimated intraclass correlation coefficient for the random intercepts was .05.

@@

In the current paper, we presented the GLMM tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters. Therefore, the GLMM tree algorithm seems preeminently suited for the detection of treatment-subgroup interactions in clustered datasets. In the empirical evaluation, we have assessed the performance of GLMM tree in a large number of simulated dataets, and compared it with GLM tree, an algorithm that creates GLM-based recursive partitions, without estimating random effects. In the application, we applied GLM and GLMM tree to an existing dataset of a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. The application showed that GLMM tree provides results that are easily interpretable, and also more accurate than a GLM tree without random effects.

The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, as it accurately recovered the interactions in 90\% of the simulated datasets with treatment-subgroup interactions. In contrast, GLM tree accurately recovered the interactions in only 61\% of the datasets with treatment-subgroup interactions. In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets. In other words, the Type I error rate of GLMM tree very closely resembled the $\alpha$ level used for evaluating significance of parameter instability, whereas the Type I error rate of GLM tree clearly exceeded this value.

@@

As expected, when random effects were absent from the simulated datasets, GLM tree and GLMM tree showed high and equal predictive accuracy. This finding indicates that GLMM tree can be applied, whenever cluster-specific random effects are expected. In the absence of random effects, GLM tree and GLMM tree are expected to perform equally well, and in the presence of random effects, GLMM tree will outperform GLM tree. This may especially be the case with large sample sizes ($N > 200$), as the increased power will likely cause GLM tree to create spurious splits, when random effects are present in the data.

@@

These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures. However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified. That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups. However, misspecification of the model can reduce performance. One source of misspecification would be, when relevant variables are not included in the GLM or as partitioning variables. If there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using the partitioning variables that are available. Or, if the coefficients of other variables vary across subgroups, then those variables should also be included in the linear predictor of the GLM. Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups. However, it should be noted that in our simulations, the number of partitioning variables did not substantially influence performance of the algorithm(s).

In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects. Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials, individual-level patient data meta-analyses, and longitudinal studies.

>>>>

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{afterpage}
\linespread{1.2}
\usepackage{tabto}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions in datasets from randomized clinical trials. In many instances, however, datasets may have a clustered structure, where observations are nested within, for example, research centers, studies or subjects. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB without random-effects estimation. GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (in 90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB without random effects (average predictive accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

In research on the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified (or personalized) medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. Often, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized. This may hamper identification of which treatment works best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate instead of test such hypotheses.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms for the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes. As such, the \citeA{ZhanyTsiaStat} method also provides a tree-based method for treatment-subgroup interaction detection.

Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection, while also taking account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models.

The new algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}). MOB offers a flexible framework for subgroup detection, and when the partitioning is based on a generalized linear model (GLM), it allows for detection of treatment-subgroup interactions. Earlier, GLM-based MOB has been applied to detect treatment-subgroup interactions in the treatment of depression \citeA{DrieySmit14} and amyotrophic lateral sclerosis \cite<ALS;>{SeibyZeil15}. The new algorithm we present combines MOB with random-effects estimates and therefore accounts for the clustered structure \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we will discuss existing frameworks the new algorithm builds on: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Application section, we provide an illustration by applying the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}. But first, we introduce an artificial motivating data set with which will be used to illustrate the methods and algorithms.

@@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM is often used to estimate treatment effects. GLMs allow for the choice of a suitable response distribution -- for example normal, binomial, or Poisson -- depending on whether the treatment outcome variable is continuous (e.g., an improvement score), binary (e.g., improved or not), or a count (e.g., number of events in a certain time span), respectively. In all cases the expectation $\mu_i$ of the outcome variable $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and a suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function\footnote{An overview of notation used is provided in the appendix.}. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. In case of a continuous response variable, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials or research centers), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

@@

In RCTs, ANCOVAs are often used to estimate average treatement effects. In these ANCOVA, treatment effects are often estimated after controlling the posttreatment values on the outcome measure for the linear effect of pretreatment values on the same measure. Therefore, we estimated the GLM and GLMM trees using posttreatment HAM-D scores, controlled for the linear effects of pretreatment HAM-D scores.

We built all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score. As a result, data from 694 patients from 7 studies were included in the analyses. Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.

To provide a standardized estimate of the treatment effect differences in the final nodes of the trees, we calculated node-specific Cohen's $d$ values. Cohen's $d$ was calculated by dividing the node-specific predicted treatment outcome difference by the node-specific pooled standard deviation. Confidence intervals for Cohen's $d$ were not calculated, as these would not take into account the exploratory nature of our analyses (i.e., variable and split point selection). The predictive accuracy of GLM and GLMM trees was assessed by calculating average correlations between observed and predicted HAM-D posttreatment scores, based on 50-fold cross validation.

<<depression-trees>>=
## offset
metadata$HRSDfit <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))

## GLM tree
lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, offset = HRSDfit)

## GLMM tree
lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + offset(HRSDfit) | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, ranefstart = metadata$HRSDfit)
@

@@

Both tree-growing algorithms were applied to the dataset; the resulting GLM tree is presented in Figure~\ref{fig:lmtree_C&W} and Table~\ref{tab:lmtree_C&W} and the resulting GLMM tree is presented in Figure~\ref{fig:lmertree_C&W} and Table~\ref{tab:lmertree_C&W}.

@@

\begin{table}[]
	\caption{Node specific summary statistics and effect sizes of HAM-D scores for GLM tree}
	\begin{tabular}{lcccc}
		\thickline
		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
		\hline
		2		&	13.25		& 10.29		& 7.52		& 0.39			\\
		4		&	7.82		& 7.55		& 5.51		& 0.05			\\
		5		&	10.35		& 7.69		& 6.64		& 0.40			\\
		\hline
		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
	\end{tabular}
	\label{tab:lmtree_C&W}
\end{table}

@@

\begin{table}[]
	\caption{Node specific summary statistics and effect size for GLMM tree}
	\label{tab:lmertree_C&W}
	\begin{tabular}{lcccc}
	\thickline
		Node	&	$\hat{\mu}_{CBT}$	& $\hat{\mu}_{PHA}$	& pooled SD	& Cohen's $d$	\\
		\hline
		2		&	8.27		& 7.99		& 5.86		& 0.05			\\
		3		&	10.84		& 8.21		& 6.80		& 0.39			\\
		\hline
		\multicolumn{5}{l}{\small \textit{Note.} $\hat{\mu}$ values are predicted posttreatment HAM-D}\\
		\multicolumn{5}{l}{\small means, evaluated at the global HAM-D pretreatment}\\
		\multicolumn{5}{l}{\small mean. Pooled SDs are calculated based on post-}\\
		\multicolumn{5}{l}{\small treatment HAM-D scores.}\\ 
	\end{tabular}
\end{table}

By taking into account study-specific intercepts, the final GLMM tree (Figure~\ref{fig:lmertree_C&W}) suggests that the first split made by GLM tree is a spurious split. The GLMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable. The estimated intraclass correlation coefficient for GLMM tree was .05.

The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$; Table~\ref{tab:lmertree_C&W}). However, for patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$; Table~\ref{tab:lmertree_C&W}).

@@

In the current paper, we presented the GLMM tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters. Therefore, the GLMM tree algorithm seems well suited for the detection of treatment-subgroup interactions in clustered datasets. This was confirmed by our simulation study, in which we compared the performance of GLMM trees with that of GLM trees without random effects. GLMM tree performed very well in recovering treatment-subgroup interactions, accurately recovering the interactions in 90\% of the simulated datasets with treatment-subgroup interactions. In contrast, GLM tree accurately recovered the interactions in only 61\% of the datasets with treatment-subgroup interactions. In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets. In other words, the Type I error rate of GLMM tree very closely approximated the $\alpha$ level used for evaluating significance of parameter instability, whereas the Type I error rate of GLM tree clearly exceeded this value.

@@

As expected, when random effects were absent from the simulated datasets, GLM tree and GLMM tree showed high and equal predictive accuracy. This is an important finding, as it indicates that GLMM tree can be applied whenever cluster-specific random effects are expected. In the absence of random effects, GLM tree and GLMM tree are expected to perform equally well, and in the presence of random effects, GLMM tree will outperform GLM tree. This may especially be the case with large sample sizes ($N > 200$), as increased power will more likely result in spurious splits, when random effects are not taken into account.

@@

In the application, we applied GLM and GLMM tree to an existing dataset of a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. This showed that GLMM tree provides results that are easily interpretable, and also more accurate than a GLM tree without random effects. We have shown how node-specific means and variances can be used to calculate effect sizes, to judge clinical relevance of the findings, as would often be done in RCTs or meta-analysis. We have limited ourselves to calculating Cohen's $d$. Although equivalent values of the success rate difference or the number needed to treat can be calculated, it would involve additional distributional assumptions \cite{KraeyKupf06}. The node-specific effect sizes can also be used to prune the tree, when a researcher prefers to have a final tree which is not only based on statistical, but also on clinical relevance. A topic for further research would be the development of splitting procedures based on effect sizes, as this would allow for taking into account clinical significant in the tree-growing process, but this is currently not possible.

These findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures. However, it should be noted that the simulations show that GLMM tree performs very well, if the model is correctly specified. That is, if there are subgroups with respect to the partitioning variables, so that there are different parameters of the GLM in each of these subgroups, then GLMM tree will accurately recover those subgroups. However, misspecification of the model can reduce performance. One source of misspecification would be the omission of relevant variables in the GLM, or as potential partitioning variables. For example, if there are actual subgroups, but the variables describing them are not entered as partitioning variables, the algorithm can only approximate the subgroups using variables that are available for partitioning. Another source of misspecification would be the inclusion of irrelevant variables, either in the linear predictor of the GLM or as partitioning variables, which may reduce the power to detect the actual subgroups. It should be noted, however, that the number of partitioning variables did not substantially influence performance of the algorithm(s) in our simulations.

In conclusion, GLMM tree provided highly accurate recovery of treatment-subgroup interactions and predictions of treatment effect differences, both in the presence and absence of cluster-specific random effects. Therefore, GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in datasets with a clustered structure, like for example in multi-center trials or individual-level patient data meta-analyses.
